---
title: "Hurricane Economic Propagation in Florida"
author: "Pietro Fraccaroli, Gimelgo Xirinda"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
---

## Libraries

```{r libraries}
# Spatial
library(sf)         # simple features: read/manipulate shapefiles
library(spdep)      # spatial dependence: contiguity matrix, spatial lags
library(tigris)     # download Census TIGER/Line shapefiles (counties, etc.)
library(blackmarbler)
library(tidyverse)
library(lubridate)
library(plm)        # fixed-effects panel models
library(glasso)
library(huge)       # optional: BIC/EBIC tuning for rho selection
library(ggplot2)
library(viridis) 
library(geojsonio)
```

---

## Florida Counties

```{r florida-counties}
# Download Florida county boundaries from Census TIGER/Line
# tigris caches the shapefile after first download
options(tigris_use_cache = TRUE)

florida <- counties(state = "FL", cb = TRUE, year = 2020) |>
  st_transform(crs = 4326)   # WGS84

# Quick check
cat("Number of counties:", nrow(florida), "\n")
cat("Columns:", paste(names(florida), collapse = ", "), "\n")

# Preview map
ggplot(florida) +
  geom_sf(fill = "lightblue", color = "white", linewidth = 0.3) +
  labs(title = "Florida Counties (n = 67)") +
  theme_minimal()
```

```{r florida-counties-df}
# Compute county centroids and build a clean reference dataframe
florida_counties <- florida |>
  st_centroid() |>
  mutate(
    lon = st_coordinates(geometry)[, 1],
    lat = st_coordinates(geometry)[, 2]
  ) |>
  st_drop_geometry() |>
  select(GEOID, NAME, lon, lat) |>
  arrange(NAME)

head(florida_counties)
```
```{r}
write.csv(florida_counties,
          file = "C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Processed/florida_counties.csv",
          row.names = FALSE)
```
```{r}
florida_counties <- read.csv("C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Processed/florida_counties.csv")
head(florida_counties)
```
## Nighttime Lights (VIIRS — Google Earth Engine)

Date range: Jan 2013 – Dec 2023 (132 months). Covers all three hurricanes
(Irma 2017, Michael 2018, Ian 2022) with ample pre/post baseline, and gives
a T=132 >> p=67 ratio needed for reliable precision matrix estimation in Stage 2.

We use `rgee` (R interface to Google Earth Engine) instead of `blackmarbler`
to avoid GDAL HDF5 driver issues on Windows. GEE runs the spatial aggregation
server-side and returns a plain dataframe.
```{r}
nighttime_lights <- read.csv("C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Raw/florida_ntl_2013_2023.csv")
head(nighttime_lights)
```

```{r ntl-merge}
florida_data <- nighttime_lights |>
  rename(ntl_mean = mean) |>
  mutate(
    GEOID = as.character(GEOID),
    date  = as.Date(date)
  ) |>
  left_join(florida_counties |> mutate(GEOID = as.character(GEOID)),
            by = "GEOID") |>
  select(GEOID, NAME = NAME.x, date, ntl_mean, lon, lat) |>
  arrange(GEOID, date)

cat("Rows:", nrow(florida_data), "— expected", 67 * 132, "\n")
head(florida_data)
```

```{r ntl-map}
# Pick one month to visualize
snapshot <- florida_data |> filter(date == as.Date("2019-06-01"))

# Join ntl values back to sf geometry for plotting
florida_ntl <- florida |>
  left_join(snapshot, by = "GEOID")

ggplot(florida_ntl) +
  geom_sf(aes(fill = ntl_mean), color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(option = "inferno", name = "Radiance\n(nW/cm²/sr)") +
  labs(title = "Nighttime Lights — Florida Counties",
       subtitle = "June 2019") +
  theme_void()
```

```{r}
summary(florida_data$ntl_mean)
```
```{r}
write.csv(florida_data,
          file = "C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Processed/florida_data.csv",
          row.names = FALSE)
```
```{r}
florida_data <- read.csv("C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Processed/florida_data.csv")
head(florida_data)
```

## Hurricane Data

```{r hurricane-load}
hurricane_data <- read.csv("C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Raw/DisasterDeclarationsSummaries.csv")
```

```{r hurricane-summary}
# All Florida hurricanes without IA filter — for reference
hurricane_data |>
  filter(state == "FL", incidentType == "Hurricane") |>
  group_by(disasterNumber, declarationTitle, incidentBeginDate) |>
  summarise(n_counties = n(),
            ia = max(iaProgramDeclared),
            pa = max(paProgramDeclared),
            .groups = "drop") |>
  arrange(incidentBeginDate) |>
  print(n = 50)
```
```{r fema-indicator}
fema_hit <- hurricane_data |>
  filter(
    state          == "FL",
    incidentType   == "Hurricane",
    fipsCountyCode != 0,
    grepl("HURRICANE", declarationTitle, ignore.case = TRUE)
  ) |>
  mutate(
    GEOID = paste0("12", sprintf("%03d", fipsCountyCode)),
    date  = as.Date(format(as.Date(substr(incidentBeginDate, 1, 10)), "%Y-%m-01"))
  ) |>
  filter(date >= as.Date("2014-01-01"), date <= as.Date("2023-12-31")) |>
  select(GEOID, date, disasterNumber, declarationTitle) |>
  distinct(GEOID, date, .keep_all = TRUE) |>
  mutate(hurricane_hit = 1L)

cat("FEMA hurricane-hit records:", nrow(fema_hit), "\n")
fema_hit |>
  group_by(disasterNumber, declarationTitle) |>
  summarise(n_counties = n(), .groups = "drop") |>
  arrange(disasterNumber) |>
  print()
```

```{r hurricane-panel}
florida_data <- read.csv(paste0(
  "C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/",
  "Data/Processed/florida_data.csv"
))

florida_data <- florida_data |>
  mutate(
    GEOID = as.character(GEOID),
    date  = as.Date(date)
  ) |>
  left_join(
    fema_hit |> mutate(GEOID = as.character(GEOID)) |>
      select(GEOID, date, hurricane_hit),
    by = c("GEOID", "date")
  ) |>
  mutate(hurricane_hit = replace_na(hurricane_hit, 0L))

cat("Panel rows:", nrow(florida_data), "— expected", 67 * 120, "\n")
cat("hurricane_hit == 1:", sum(florida_data$hurricane_hit), "\n")
head(florida_data)
```

```{r hurricane-maps}
florida <- counties(state = "FL", cb = TRUE, year = 2020) |>
  st_transform(crs = 4326)

fig_dir <- paste0(
  "C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Figures/"
)

hurricanes <- fema_hit |>
  distinct(disasterNumber, declarationTitle, date)

for (i in seq_len(nrow(hurricanes))) {
  dis_num  <- hurricanes$disasterNumber[i]
  title    <- hurricanes$declarationTitle[i]
  dis_date <- hurricanes$date[i]

  hit_geoids <- fema_hit |>
    filter(disasterNumber == dis_num) |>
    pull(GEOID)

  map_data <- florida |>
    mutate(
      GEOID = as.character(GEOID),
      hit   = GEOID %in% hit_geoids
    )

  p <- ggplot(map_data) +
    geom_sf(aes(fill = hit), color = "white", linewidth = 0.2) +
    scale_fill_manual(values = c("FALSE" = "lightgrey", "TRUE" = "red"),
                      guide = "none") +
    labs(title = title,
         subtitle = format(dis_date, "%B %Y")) +
    theme_void()

  fname <- paste0(fig_dir, "hurricane_", dis_num, ".png")
  ggsave(fname, plot = p, width = 6, height = 5, dpi = 150)
  cat("Saved:", fname, "\n")
}
```