---
title: "Hurricane Economic Propagation in Florida"
author: "Pietro Fraccaroli, Gimelgo Xirinda"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
---

## Libraries

```{r libraries}
# Spatial
library(sf)         # simple features: read/manipulate shapefiles
library(spdep)      # spatial dependence: contiguity matrix, spatial lags
library(tigris)     # download Census TIGER/Line shapefiles (counties, etc.)
library(blackmarbler)
library(tidyverse)
library(lubridate)
library(plm)        # fixed-effects panel models
library(glasso)
library(huge)       # optional: BIC/EBIC tuning for rho selection
library(ggplot2)
library(viridis) 
library(geojsonio)
```

---

## Florida Counties

```{r florida-counties}
# Download Florida county boundaries from Census TIGER/Line
# tigris caches the shapefile after first download
options(tigris_use_cache = TRUE)

florida <- counties(state = "FL", cb = TRUE, year = 2020) |>
  st_transform(crs = 4326)   # WGS84

# Quick check
cat("Number of counties:", nrow(florida), "\n")
cat("Columns:", paste(names(florida), collapse = ", "), "\n")

# Preview map
ggplot(florida) +
  geom_sf(fill = "lightblue", color = "white", linewidth = 0.3) +
  labs(title = "Florida Counties (n = 67)") +
  theme_minimal()
```

```{r florida-counties-df}
# Compute county centroids and build a clean reference dataframe
florida_counties <- florida |>
  st_centroid() |>
  mutate(
    lon = st_coordinates(geometry)[, 1],
    lat = st_coordinates(geometry)[, 2]
  ) |>
  st_drop_geometry() |>
  select(GEOID, NAME, lon, lat) |>
  arrange(NAME)

head(florida_counties)
```
```{r}
write.csv(florida_counties,
          file = "C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Processed/florida_counties.csv",
          row.names = FALSE)
```
```{r}
florida_counties <- read.csv("C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Processed/florida_counties.csv")
head(florida_counties)
```
## Nighttime Lights (VIIRS — Google Earth Engine)

Date range: Jan 2013 – Dec 2023 (132 months). Covers all three hurricanes
(Irma 2017, Michael 2018, Ian 2022) with ample pre/post baseline, and gives
a T=132 >> p=67 ratio needed for reliable precision matrix estimation in Stage 2.

We use `rgee` (R interface to Google Earth Engine) instead of `blackmarbler`
to avoid GDAL HDF5 driver issues on Windows. GEE runs the spatial aggregation
server-side and returns a plain dataframe.
```{r}
nighttime_lights <- read.csv("C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Raw/florida_ntl_2013_2023.csv")
head(nighttime_lights)
```

```{r ntl-merge}
florida_data <- nighttime_lights |>
  rename(ntl_mean = mean) |>
  mutate(
    GEOID = as.character(GEOID),
    date  = as.Date(date)
  ) |>
  left_join(florida_counties |> mutate(GEOID = as.character(GEOID)),
            by = "GEOID") |>
  select(GEOID, NAME = NAME.x, date, ntl_mean, lon, lat) |>
  arrange(GEOID, date)

cat("Rows:", nrow(florida_data), "— expected", 67 * 132, "\n")
head(florida_data)
```

```{r ntl-map}
# Pick one month to visualize
snapshot <- florida_data |> filter(date == as.Date("2019-06-01"))

# Join ntl values back to sf geometry for plotting
florida_ntl <- florida |>
  left_join(snapshot, by = "GEOID")

ggplot(florida_ntl) +
  geom_sf(aes(fill = ntl_mean), color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(option = "inferno", name = "Radiance\n(nW/cm²/sr)") +
  labs(title = "Nighttime Lights — Florida Counties",
       subtitle = "June 2019") +
  theme_void()
```

```{r}
summary(florida_data$ntl_mean)
```
```{r}
write.csv(florida_data,
          file = "C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Processed/florida_data.csv",
          row.names = FALSE)
```
```{r}
florida_data <- read.csv("C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Processed/florida_data.csv")
head(florida_data)
```
## Hurricane Data

```{r hurricane-load}
hurricane_data <- read.csv("C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Raw/DisasterDeclarationsSummaries.csv")
```

```{r}
# Tutti gli uragani in Florida senza filtro IA
hurricane_data |>
  filter(state == "FL", incidentType == "Hurricane") |>
  group_by(disasterNumber, declarationTitle, incidentBeginDate) |>
  summarise(n_counties = n(),
            ia = max(iaProgramDeclared),
            pa = max(paProgramDeclared),
            .groups = "drop") |>
  arrange(incidentBeginDate) |>
  print(n = 50)

```