---
title: "Hurricane Economic Propagation in Florida"
author: "Pietro Fraccaroli, Gimelgo Xirinda"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
---

## Libraries

```{r libraries}
# Spatial
library(sf)         # simple features: read/manipulate shapefiles
library(spdep)      # spatial dependence: contiguity matrix, spatial lags
library(tigris)     # download Census TIGER/Line shapefiles (counties, etc.)
library(blackmarbler)
library(tidyverse)
library(lubridate)
library(plm)        # fixed-effects panel models
library(glasso)
library(huge)       # optional: BIC/EBIC tuning for rho selection
library(ggplot2)
library(viridis) 
library(geojsonio)
```

---

## Florida Counties

```{r florida-counties}
# Download Florida county boundaries from Census TIGER/Line
# tigris caches the shapefile after first download
options(tigris_use_cache = TRUE)

florida <- counties(state = "FL", cb = TRUE, year = 2020) |>
  st_transform(crs = 4326)   # WGS84

# Quick check
cat("Number of counties:", nrow(florida), "\n")
cat("Columns:", paste(names(florida), collapse = ", "), "\n")

# Preview map
ggplot(florida) +
  geom_sf(fill = "lightblue", color = "white", linewidth = 0.3) +
  labs(title = "Florida Counties (n = 67)") +
  theme_minimal()
```

```{r florida-counties-df}
# Compute county centroids and build a clean reference dataframe
florida_counties <- florida |>
  st_centroid() |>
  mutate(
    lon = st_coordinates(geometry)[, 1],
    lat = st_coordinates(geometry)[, 2]
  ) |>
  st_drop_geometry() |>
  select(GEOID, NAME, lon, lat) |>
  arrange(NAME)

head(florida_counties)
```
```{r}
write.csv(florida_counties,
          file = "C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Processed/florida_counties.csv",
          row.names = FALSE)
```
```{r}
florida_counties <- read.csv("C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Processed/florida_counties.csv")
head(florida_counties)
```
## Nighttime Lights (VIIRS — Google Earth Engine)

Date range: Jan 2013 – Dec 2023 (132 months). Covers all three hurricanes
(Irma 2017, Michael 2018, Ian 2022) with ample pre/post baseline, and gives
a T=132 >> p=67 ratio needed for reliable precision matrix estimation in Stage 2.

We use `rgee` (R interface to Google Earth Engine) instead of `blackmarbler`
to avoid GDAL HDF5 driver issues on Windows. GEE runs the spatial aggregation
server-side and returns a plain dataframe.

**One-time setup:** `install.packages("rgee")`, then in the console run
`reticulate::py_install("earthengine-api", pip = TRUE)` to install the
Python dependency into the existing env (avoids `ee_install()` conflicts
with uv-managed Python on Windows).

```{r ntl-initialize}
# Point reticulate at the correct Python before loading rgee
reticulate::use_condaenv("r-reticulate", required = TRUE)
library(rgee)
```
```{r}
ee_Authenticate()  
```
```{r}
ee_Initialize()
```

```{r ntl-download}
florida_ee <- sf_as_ee(florida |> select(GEOID, NAME))

viirs <- ee$ImageCollection("NOAA/VIIRS/DNB/MONTHLY_V1/VCMSLCFG") |>
  ee$ImageCollection$filterDate("2013-01-01", "2023-12-31") |>
  ee$ImageCollection$select("avg_rad")

# reduceRegions called on image object directly (compatible with new API)
extract_month <- function(img) {
  img     <- ee$Image(img)
  date    <- img$date()$format("YYYY-MM-dd")
  reduced <- img$reduceRegions(
    collection = florida_ee,
    reducer    = ee$Reducer$mean(),
    scale      = 500
  )
  reduced$map(function(f) f$set("date", date))
}

results <- viirs$map(extract_month)$flatten()
ntl_raw <- ee_as_sf(results, via = "drive") |> st_drop_geometry()

cat("Dimensions:", nrow(ntl_raw), "x", ncol(ntl_raw), "\n")
print(head(ntl_raw))
```

```{r ntl-load}
ntl_raw <- readRDS("../Data/Processed/ntl_raw.rds")

# ee_extract returns one column per image named by date (e.g. X202201)
# Pivot to long format: one row per county x month
ntl <- ntl_raw |>
  pivot_longer(
    cols      = starts_with("X"),
    names_to  = "date",
    values_to = "ntl_mean"
  ) |>
  mutate(date = as.Date(paste0(sub("X", "", date), "01"), "%Y%m%d")) |>
  rename(GEOID = GEOID, county = NAME) |>
  arrange(GEOID, date)

glimpse(ntl)
cat("\nRows:", nrow(ntl),
    "— expected 67 x 132 =", 67 * 132, "\n")
print(head(ntl, 10))
```
