---
title: "Hurricane Economic Propagation in Florida"
author: "Pietro Fraccaroli, Gimelgo Xirinda"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
---

## Libraries

```{r libraries}
# Spatial
library(sf)         # simple features: read/manipulate shapefiles
library(spdep)      # spatial dependence: contiguity matrix, spatial lags
library(tigris)     # download Census TIGER/Line shapefiles (counties, etc.)
library(blackmarbler)
library(tidyverse)
library(lubridate)
library(plm)        # fixed-effects panel models
library(glasso)
library(huge)       # optional: BIC/EBIC tuning for rho selection
library(ggplot2)
library(viridis) 
library(geojsonio)
```

---

## Florida Counties

```{r florida-counties}
# Download Florida county boundaries from Census TIGER/Line
# tigris caches the shapefile after first download
options(tigris_use_cache = TRUE)

florida <- counties(state = "FL", cb = TRUE, year = 2020) |>
  st_transform(crs = 4326)   # WGS84

# Quick check
cat("Number of counties:", nrow(florida), "\n")
cat("Columns:", paste(names(florida), collapse = ", "), "\n")

# Preview map
ggplot(florida) +
  geom_sf(fill = "lightblue", color = "white", linewidth = 0.3) +
  labs(title = "Florida Counties (n = 67)") +
  theme_minimal()
```

```{r florida-counties-df}
# Compute county centroids and build a clean reference dataframe
florida_counties <- florida |>
  st_centroid() |>
  mutate(
    lon = st_coordinates(geometry)[, 1],
    lat = st_coordinates(geometry)[, 2]
  ) |>
  st_drop_geometry() |>
  select(GEOID, NAME, lon, lat) |>
  arrange(NAME)

head(florida_counties)
```
```{r}
write.csv(florida_counties,
          file = "C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Processed/florida_counties.csv",
          row.names = FALSE)
```
```{r}
florida_counties <- read.csv("C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Processed/florida_counties.csv")
head(florida_counties)
```
## Nighttime Lights (VIIRS — Google Earth Engine)

Date range: Jan 2013 – Dec 2023 (132 months). Covers all three hurricanes
(Irma 2017, Michael 2018, Ian 2022) with ample pre/post baseline, and gives
a T=132 >> p=67 ratio needed for reliable precision matrix estimation in Stage 2.

We use `rgee` (R interface to Google Earth Engine) instead of `blackmarbler`
to avoid GDAL HDF5 driver issues on Windows. GEE runs the spatial aggregation
server-side and returns a plain dataframe.
```{r}
nighttime_lights <- read.csv("C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Raw/florida_ntl_2013_2023.csv")
head(nighttime_lights)
```

```{r ntl-merge}
florida_data <- nighttime_lights |>
  rename(ntl_mean = mean) |>
  mutate(
    GEOID = as.character(GEOID),
    date  = as.Date(date)
  ) |>
  left_join(florida_counties |> mutate(GEOID = as.character(GEOID)),
            by = "GEOID") |>
  select(GEOID, NAME = NAME.x, date, ntl_mean, lon, lat) |>
  arrange(GEOID, date)

cat("Rows:", nrow(florida_data), "— expected", 67 * 132, "\n")
head(florida_data)
```

```{r ntl-map}
# Pick one month to visualize
snapshot <- florida_data |> filter(date == as.Date("2019-06-01"))

# Join ntl values back to sf geometry for plotting
florida_ntl <- florida |>
  left_join(snapshot, by = "GEOID")

ggplot(florida_ntl) +
  geom_sf(aes(fill = ntl_mean), color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(option = "inferno", name = "Radiance\n(nW/cm²/sr)") +
  labs(title = "Nighttime Lights — Florida Counties",
       subtitle = "June 2019") +
  theme_void()
```

```{r}
summary(florida_data$ntl_mean)
```
```{r}
write.csv(florida_data,
          file = "C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Processed/florida_data.csv",
          row.names = FALSE)
```
```{r}
florida_data <- read.csv("C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Processed/florida_data.csv")
head(florida_data)
```

## Hurricane Data

```{r hurricane-load}
hurricane_data <- read.csv("C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Data/Raw/DisasterDeclarationsSummaries.csv")
```

```{r hurricane-summary}
# All Florida hurricanes without IA filter — for reference
hurricane_data |>
  filter(state == "FL", incidentType == "Hurricane") |>
  group_by(disasterNumber, declarationTitle, incidentBeginDate) |>
  summarise(n_counties = n(),
            ia = max(iaProgramDeclared),
            pa = max(paProgramDeclared),
            .groups = "drop") |>
  arrange(incidentBeginDate) |>
  print(n = 50)
```
```{r}
hurricane_data |>
  filter(state == "FL", incidentType == "Hurricane") |>
  distinct(disasterNumber) |>
  nrow()

```
```{r fema-indicator}
fema_hit <- hurricane_data |>
  filter(
    state          == "FL",
    incidentType   == "Hurricane",
    fipsCountyCode != 0,
    disasterNumber >= 4000,
    grepl("HURRICANE", declarationTitle, ignore.case = TRUE)
  ) |>
  mutate(
    GEOID = paste0("12", sprintf("%03d", fipsCountyCode)),
    date  = as.Date(format(as.Date(substr(incidentBeginDate, 1, 10)), "%Y-%m-01"))
  ) |>
  filter(date >= as.Date("2014-01-01"), date <= as.Date("2023-12-31")) |>
  select(GEOID, date, disasterNumber, declarationTitle) |>
  distinct(GEOID, date, .keep_all = TRUE) |>
  mutate(hurricane_hit = 1L)

cat("FEMA hurricane-hit records:", nrow(fema_hit), "\n")
fema_hit |>
  group_by(disasterNumber, declarationTitle) |>
  summarise(n_counties = n(), .groups = "drop") |>
  arrange(disasterNumber) |>
  print()
```

```{r hurricane-panel}
florida_data <- read.csv(paste0(
  "C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/",
  "Data/Processed/florida_data.csv"
))

florida_data <- florida_data |>
  mutate(
    GEOID = as.character(GEOID),
    date  = as.Date(date)
  ) |>
  left_join(
    fema_hit |> mutate(GEOID = as.character(GEOID)) |>
      select(GEOID, date, hurricane_hit),
    by = c("GEOID", "date")
  ) |>
  mutate(hurricane_hit = replace_na(hurricane_hit, 0L))

cat("Panel rows:", nrow(florida_data), "— expected", 67 * 120, "\n")
cat("hurricane_hit == 1:", sum(florida_data$hurricane_hit), "\n")
head(florida_data)

write.csv(florida_data,
          file = paste0(
            "C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/",
            "Data/Processed/florida_data.csv"
          ),
          row.names = FALSE)
```

```{r hurricane-maps}
florida <- counties(state = "FL", cb = TRUE, year = 2020) |>
  st_transform(crs = 4326)

fig_dir <- paste0(
  "C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/Figures/"
)

hurricanes <- fema_hit |>
  distinct(disasterNumber, declarationTitle, date)

for (i in seq_len(nrow(hurricanes))) {
  dis_num  <- hurricanes$disasterNumber[i]
  title    <- hurricanes$declarationTitle[i]
  dis_date <- hurricanes$date[i]

  hit_geoids <- fema_hit |>
    filter(disasterNumber == dis_num) |>
    pull(GEOID)

  map_data <- florida |>
    mutate(
      GEOID = as.character(GEOID),
      hit   = GEOID %in% hit_geoids
    )

  p <- ggplot(map_data) +
    geom_sf(aes(fill = hit), color = "white", linewidth = 0.2) +
    scale_fill_manual(values = c("FALSE" = "lightgrey", "TRUE" = "red"),
                      guide = "none") +
    labs(title = title,
         subtitle = format(dis_date, "%B %Y")) +
    theme_void()

  fname <- paste0(fig_dir, "hurricane_", dis_num, ".png")
  ggsave(fname, plot = p, width = 6, height = 5, dpi = 150)
  cat("Saved:", fname, "\n")
}
```

---

## Stage 1 — Panel Regression

### Contiguity Matrix

We are adding a contiguity matrix to capture the spatial spillover. If we omit this we find banal connection between confinant counties.

```{r contiguity}
florida <- counties(state = "FL", cb = TRUE, year = 2020) |>
  st_transform(crs = 4326)

# Rook contiguity (shared border only)
nb <- poly2nb(florida, queen = FALSE)
W  <- nb2mat(nb, style = "B", zero.policy = TRUE)

# Rows/cols ordered by florida$GEOID
rownames(W) <- florida$GEOID
colnames(W) <- florida$GEOID

cat("W dimensions:", dim(W), "\n")
cat("Avg neighbours per county:", mean(rowSums(W)), "\n")
```

### Spatial Lag

```{r spatial-lag}
florida_data <- read.csv(paste0(
  "C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/",
  "Data/Processed/florida_data.csv"
))

florida_data <- florida_data |>
  mutate(GEOID = as.character(GEOID), date = as.Date(date))

# Wide matrix: rows = months, cols = counties (ordered as in W)
county_order <- colnames(W)
dates        <- sort(unique(florida_data$date))

hit_matrix <- florida_data |>
  select(GEOID, date, hurricane_hit) |>
  pivot_wider(names_from = GEOID, values_from = hurricane_hit) |>
  arrange(date) |>
  select(all_of(county_order)) |>
  as.matrix()

# Spatial lag: for each month, W %*% hit_vector
lag_matrix <- hit_matrix %*% t(W)

# Back to long format
lag_long <- as.data.frame(lag_matrix) |>
  mutate(date = dates) |>
  pivot_longer(-date, names_to = "GEOID", values_to = "hurricane_lag")

florida_data <- florida_data |>
  left_join(lag_long, by = c("GEOID", "date"))

cat("Spatial lag range:", range(florida_data$hurricane_lag), "\n")
head(florida_data)
```
```{r}
View(florida_data)
```
### Panel Regression

```{r panel-regression}
pdata <- pdata.frame(florida_data, index = c("GEOID", "date"))

model <- plm(
  ntl_mean ~ hurricane_hit + hurricane_lag,
  data   = pdata,
  effect = "twoways",
  model  = "within"
)

summary(model)

florida_data$residuals <- as.numeric(residuals(model))
cat("Residuals range:", range(florida_data$residuals), "\n")
```

---

## Stage 2 — Graphical Lasso on Residuals

First matrix with residuals, composed by 120 rows (months) and 67 columns (counties).
```{r glasso-matrix}
# Build T x 67 residual matrix (rows = months, cols = counties)
county_order <- colnames(W)   # same ordering as contiguity matrix
dates        <- sort(unique(florida_data$date))

resid_matrix <- florida_data |>
  select(GEOID, date, residuals) |>
  pivot_wider(names_from = GEOID, values_from = residuals) |>
  arrange(date) |>
  select(all_of(county_order)) |>
  as.matrix()

cat("Residual matrix dimensions:", dim(resid_matrix), "\n")   # 120 x 67
cat("Any NA:", anyNA(resid_matrix), "\n")
```

Here we tune the glasso penalty parameter rho using EBIC using the huge package, which evaluates a grid of rho values and selects the one that minimizes EBIC.
```{r glasso-tune}
# Estimate sample covariance matrix (67 x 67)
S <- cov(resid_matrix)

# Tune rho via EBIC (Extended BIC) using the huge package
# huge.select returns the rho that minimises EBIC over a grid
huge_out    <- huge(resid_matrix, method = "glasso", nlambda = 30, verbose = FALSE)
huge_sel    <- huge.select(huge_out, criterion = "ebic", verbose = FALSE)
rho_optimal <- huge_sel$opt.lambda

cat("Optimal rho (EBIC):", round(rho_optimal, 4), "\n")
```

Optimal rho (EBIC): 0.0996 
Here we fit the glasso with the selected rho, where it estimates the sparse precision matrix Theta. 
We shrink small values to zero.
```{r glasso-fit}
# Fit graphical lasso at the selected rho
fit   <- glasso(S, rho = rho_optimal)
Theta <- fit$wi   # 67 x 67 precision matrix

# Adjacency matrix: 1 where Theta[i,j] != 0 (off-diagonal)
adj_glasso <- (abs(Theta) > 1e-6) * 1L
diag(adj_glasso) <- 0L
rownames(adj_glasso) <- county_order
colnames(adj_glasso) <- county_order

cat("Edges in glasso network:", sum(adj_glasso) / 2, "\n")
cat("Edges in contiguity W:  ", sum(W) / 2, "\n")
```

```{r glasso-compare}
# Reorder W to match adj_glasso (same county_order)
W_ordered <- W[county_order, county_order]

# Edges in glasso but NOT in contiguity = candidate non-geographic channels
non_geo  <- adj_glasso * (1L - W_ordered)
geo_only <- W_ordered  * (1L - adj_glasso)
both     <- adj_glasso * W_ordered

cat("Edges in both (geo + economic):", sum(both) / 2, "\n")
cat("Edges only in glasso (non-geo):", sum(non_geo) / 2, "\n")
cat("Edges only in W (geo, not glasso):", sum(geo_only) / 2, "\n")
```

```{r glasso-map}
# Map non-geographic glasso edges as lines between county centroids
florida_counties <- read.csv(paste0(
  "C:/Users/Utente/OneDrive/Documenti/GitHub/Advanced_Economics/",
  "Data/Processed/florida_counties.csv"
)) |> mutate(GEOID = as.character(GEOID))

# Build edge list for non-geographic connections
edge_list <- which(non_geo == 1, arr.ind = TRUE)
edge_list <- edge_list[edge_list[, 1] < edge_list[, 2], ]   # upper triangle only

edges_df <- data.frame(
  from = rownames(non_geo)[edge_list[, 1]],
  to   = colnames(non_geo)[edge_list[, 2]]
) |>
  left_join(florida_counties |> select(GEOID, lon, lat),
            by = c("from" = "GEOID")) |>
  rename(lon_from = lon, lat_from = lat) |>
  left_join(florida_counties |> select(GEOID, lon, lat),
            by = c("to" = "GEOID")) |>
  rename(lon_to = lon, lat_to = lat)

florida_sf <- counties(state = "FL", cb = TRUE, year = 2020) |>
  st_transform(crs = 4326)

ggplot() +
  geom_sf(data = florida_sf, fill = "grey95", color = "grey70", linewidth = 0.2) +
  geom_segment(data = edges_df,
               aes(x = lon_from, y = lat_from, xend = lon_to, yend = lat_to),
               color = "steelblue", linewidth = 0.5, alpha = 0.7) +
  labs(title = "Non-Geographic Economic Linkages (Graphical Lasso)",
       subtitle = "Edges not explained by geographic contiguity") +
  theme_void()
```

```{r glasso-degree}
# County names lookup
county_names <- florida_counties |> select(GEOID, NAME)

# Degree in the full glasso network (all 66 edges)
degree_full <- data.frame(
  GEOID  = county_order,
  degree_total = rowSums(adj_glasso)
) |> left_join(county_names, by = "GEOID")

# Degree in non-geographic network only (60 edges)
degree_nongeo <- data.frame(
  GEOID        = county_order,
  degree_nongeo = rowSums(non_geo)
) |> left_join(county_names, by = "GEOID")

degree_summary <- degree_full |>
  left_join(degree_nongeo |> select(GEOID, degree_nongeo), by = "GEOID") |>
  arrange(desc(degree_nongeo)) |>
  select(NAME, degree_total, degree_nongeo)

print(degree_summary)
```

```{r glasso-degree-map}
# Choropleth: shade each county by its non-geographic degree
florida_sf <- counties(state = "FL", cb = TRUE, year = 2020) |>
  st_transform(crs = 4326) |>
  mutate(GEOID = as.character(GEOID)) |>
  left_join(degree_nongeo |> select(GEOID, degree_nongeo), by = "GEOID")

ggplot(florida_sf) +
  geom_sf(aes(fill = degree_nongeo), color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(option = "plasma", name = "Non-geo\ndegree") +
  labs(title = "Non-Geographic Economic Connectivity (Graphical Lasso)",
       subtitle = "Number of non-contiguous counties each county is linked to") +
  theme_void()
```

This code give us a list of the strongest non-geographic connections. Considering also the partial correlation.
```{r glasso-strength}
# Partial correlation matrix from precision matrix
# partial_corr[i,j] = -Theta[i,j] / sqrt(Theta[i,i] * Theta[j,j])
D    <- diag(Theta)
Pcor <- -Theta / sqrt(outer(D, D))
diag(Pcor) <- 1

rownames(Pcor) <- county_order
colnames(Pcor) <- county_order

# Top 20 strongest non-geographic edges by absolute partial correlation
nongeo_idx <- which(non_geo == 1, arr.ind = TRUE)
nongeo_idx <- nongeo_idx[nongeo_idx[, 1] < nongeo_idx[, 2], ]

strength_df <- data.frame(
  from        = county_order[nongeo_idx[, 1]],
  to          = county_order[nongeo_idx[, 2]],
  partial_cor = Pcor[nongeo_idx]
) |>
  left_join(county_names, by = c("from" = "GEOID")) |> rename(name_from = NAME) |>
  left_join(county_names, by = c("to"   = "GEOID")) |> rename(name_to   = NAME) |>
  mutate(abs_pcor = abs(partial_cor)) |>
  arrange(desc(abs_pcor)) |>
  select(name_from, name_to, partial_cor, abs_pcor)

print(strength_df)
```

Geographic connections.
```{r glasso-geo-map}
# Map the 6 edges that are BOTH in glasso AND in W (geographic + economic)
geo_idx <- which(both == 1, arr.ind = TRUE)
geo_idx <- geo_idx[geo_idx[, 1] < geo_idx[, 2], ]

geo_edges_df <- data.frame(
  from = rownames(both)[geo_idx[, 1]],
  to   = colnames(both)[geo_idx[, 2]]
) |>
  left_join(florida_counties |> select(GEOID, lon, lat),
            by = c("from" = "GEOID")) |>
  rename(lon_from = lon, lat_from = lat) |>
  left_join(florida_counties |> select(GEOID, lon, lat),
            by = c("to" = "GEOID")) |>
  rename(lon_to = lon, lat_to = lat)

ggplot() +
  geom_sf(data = florida_sf, fill = "grey95", color = "grey70", linewidth = 0.2) +
  geom_segment(data = geo_edges_df,
               aes(x = lon_from, y = lat_from, xend = lon_to, yend = lat_to),
               color = "darkred", linewidth = 1, alpha = 0.8) +
  labs(title = "Geographic + Economic Linkages (Graphical Lasso)",
       subtitle = "Edges in both glasso network and contiguity matrix (n = 6)") +
  theme_void()
```

---

## Validazione — Residui di Orange nel Tempo

```{r orange-residuals}
hurricane_dates <- data.frame(
  date  = as.Date(c("2017-09-01", "2018-10-01", "2022-09-01")),
  label = c("Irma", "Michael", "Ian")
)

florida_data |>
  mutate(GEOID = as.character(GEOID), date = as.Date(date)) |>
  filter(NAME == "Orange") |>
  ggplot(aes(x = date, y = residuals)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(data = hurricane_dates,
             aes(xintercept = date),
             color = "red", linetype = "dashed", linewidth = 0.8) +
  geom_text(data = hurricane_dates,
            aes(x = date, y = max(florida_data$residuals[florida_data$NAME == "Orange"], na.rm = TRUE),
                label = label),
            vjust = -0.3, hjust = -0.1, color = "red", size = 3.5) +
  labs(title = "Residuals NTL — Orange County",
       subtitle = "Red lines = hurricane dates (Irma 2017, Michael 2018, Ian 2022)",
       x = NULL, y = "Residual") +
  theme_minimal()
```

---

## Validation — Extended Regression with Glasso Network Spillover

```{r glasso-spillover}
# For each county-month, count how many non-geographic glasso neighbors were hit
glasso_lag_matrix <- florida_data |>
  mutate(GEOID = as.character(GEOID), date = as.Date(date)) |>
  select(GEOID, date, hurricane_hit) |>
  pivot_wider(names_from = GEOID, values_from = hurricane_hit) |>
  arrange(date) |>
  select(all_of(county_order)) |>
  as.matrix()

glasso_lag_counts <- glasso_lag_matrix %*% non_geo

glasso_lag_long <- as.data.frame(glasso_lag_counts) |>
  mutate(date = sort(unique(as.Date(florida_data$date)))) |>
  pivot_longer(-date, names_to = "GEOID", values_to = "glasso_neighbor_hit")

florida_data2 <- florida_data |>
  mutate(GEOID = as.character(GEOID), date = as.Date(date)) |>
  left_join(glasso_lag_long, by = c("GEOID", "date"))

pdata2 <- pdata.frame(florida_data2, index = c("GEOID", "date"))

model2 <- plm(
  ntl_mean ~ hurricane_hit + hurricane_lag + glasso_neighbor_hit,
  data   = pdata2,
  effect = "twoways",
  model  = "within"
)

summary(model2)
```
